\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{physics}
\usepackage[margin=2.5cm]{geometry}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{hyperref}

% Code listing style for Python
\lstdefinestyle{python}{
    language=Python,
    basicstyle=\ttfamily\small,
    keywordstyle=\color{blue!70!black}\bfseries,
    stringstyle=\color{red!60!black},
    commentstyle=\color{green!50!black}\itshape,
    numberstyle=\tiny\color{gray},
    numbers=left,
    numbersep=8pt,
    frame=single,
    framerule=0.4pt,
    rulecolor=\color{gray!50},
    backgroundcolor=\color{gray!5},
    breaklines=true,
    showstringspaces=false,
    tabsize=4,
    xleftmargin=2em,
    framexleftmargin=1.5em,
}

\title{Missing Time-Reversal Conjugation of $W_{\vb{GG}'}(\vb{q})$
       in the BSE Direct Kernel}
\author{}
\date{}

\begin{document}
\maketitle

%----------------------------------------------------------------------
\section{Summary}
%----------------------------------------------------------------------

The Bethe--Salpeter equation (BSE) direct kernel in GPAW contracts the
screened Coulomb interaction $W_{\vb{GG}'}(\vb{q})$ with pair densities
for valence--valence and conduction--conduction transitions.  When
the Brillouin-zone (BZ) momentum transfer $\vb{Q}$ is mapped to its
irreducible representative $\vb{q}_\text{IBZ}$ via a symmetry operation
that includes time reversal, the stored $W_{\vb{GG}'}(\vb{q}_\text{IBZ})$
must be complex-conjugated before contraction.  This conjugation was
missing, leading to a non-Hermitian BSE Hamiltonian for systems lacking
inversion symmetry (e.g.\ monolayer MoS\textsubscript{2}).

%----------------------------------------------------------------------
\section{Origin of the symmetry operation}
%----------------------------------------------------------------------

GPAW computes $W_{\vb{GG}'}(\vb{q})$ only at the irreducible
$\vb{q}$-points.  When the BSE kernel loops over all $(K_1, K_2)$
pairs, the momentum transfer
\begin{equation}
  \vb{Q} = \vb{K}_2 - \vb{K}_1
\end{equation}
is a general BZ $\vb{q}$-point that may not lie in the IBZ.  The code
therefore finds a symmetry operation $\hat{S}$ (possibly combined with
time reversal $\hat{T}$) such that
\begin{equation}\label{eq:qmap}
  \vb{Q} = s\, U \,\vb{q}_\text{IBZ},
  \qquad
  s = \begin{cases} +1 & \text{(pure rotation)}, \\
                     -1 & \text{(rotation + time reversal)}.
  \end{cases}
\end{equation}
This mapping is performed in the class method
\texttt{QSymmetryOp.get\_symop\_from\_kpair}:

\begin{lstlisting}[style=python,firstnumber=263,
  caption={\texttt{gpaw/response/g0w0.py}: Determination of the
           symmetry operation mapping $\vb{Q}$ to $\vb{q}_\text{IBZ}$.}]
def get_symop_from_kpair(cls, kd, qd, kpt1, kpt2):
    Q_c = kd.bzk_kc[kpt2.K] - kd.bzk_kc[kpt1.K]
    iQ = qd.where_is_q(Q_c, qd.bzk_kc)
    iq = qd.bz2ibz_k[iQ]
    q_c = qd.ibzk_kc[iq]

    sym = qd.sym_k[iQ]
    U_cc = qd.symmetry.op_scc[sym]
    time_reversal = qd.time_reversal_k[iQ]
    sign = 1 - 2 * time_reversal
    symop = cls(sym, U_cc, sign)
    symop.check_q_Q_symmetry(Q_c, q_c)

    return symop, iq
\end{lstlisting}

The sign $s$ propagates into the pair-density calculation via the
\texttt{apply} method (line~228), which maps reciprocal-lattice vectors
as $\vb{G} \to s\,U\,\vb{G}$, and via the PAW correction
\texttt{remap} method, which conjugates the PAW tensor when $s = -1$
(line~456 of \texttt{gpaw/response/paw.py}).

%----------------------------------------------------------------------
\section{Why time reversal is needed for MoS\textsubscript{2}}
%----------------------------------------------------------------------

Monolayer MoS\textsubscript{2} has point group $D_{3h}$
($\bar{6}m2$): three-fold rotations and mirror planes, but
\emph{no inversion center}.  For a centrosymmetric crystal, the
spatial inversion maps $\vb{q} \to -\vb{q}$, so every BZ
$\vb{q}$-point can be folded to the IBZ using spatial symmetry alone
($s = +1$ always).  Without inversion, some $\vb{q}$-points require
time reversal ($\vb{q} \to -\vb{q}$) to reach the IBZ, giving
$s = -1$.

%----------------------------------------------------------------------
\section{Transformation of $W_{\vb{GG}'}$ under time reversal}
%----------------------------------------------------------------------

Since $W(\vb{r}, \vb{r}', \omega = 0)$ is real in real space, its
plane-wave matrix elements satisfy
\begin{equation}\label{eq:reality}
  W_{\vb{GG}'}(\vb{q})^*
  = W_{-\vb{G},-\vb{G}'}(-\vb{q}).
\end{equation}
The contraction in the BSE kernel requires $W$ at the BZ wavevectors
$\vb{Q} + \vb{G}_\text{BZ}$.  From Eq.~\eqref{eq:qmap}, the
pair-density code maps the IBZ reciprocal-lattice vectors as
$\vb{G}_\text{BZ} = s\,U\,\vb{G}_\text{IBZ}$, so we need
$W_{\vb{G}_\text{BZ},\,\vb{G}_\text{BZ}'}(\vb{Q})$.
Rotational invariance of $W$ in reciprocal space gives
\begin{equation}\label{eq:rotation}
  W_{U\vb{G},\,U\vb{G}'}(U\vb{q})
  = W_{\vb{GG}'}(\vb{q}).
\end{equation}

For a \emph{pure rotation} ($s = +1$), $\vb{Q} = U\vb{q}_\text{IBZ}$
and $\vb{G}_\text{BZ} = U\vb{G}_\text{IBZ}$, so by
Eq.~\eqref{eq:rotation}:
\begin{equation}
  W_{U\vb{G},\,U\vb{G}'}(U\vb{q}_\text{IBZ})
  = W_{\vb{GG}'}(\vb{q}_\text{IBZ}),
\end{equation}
and the stored IBZ matrix can be used directly.

For a \emph{time-reversed} operation ($s = -1$),
$\vb{Q} = -U\vb{q}_\text{IBZ}$ and
$\vb{G}_\text{BZ} = -U\vb{G}_\text{IBZ}$.  Applying
Eq.~\eqref{eq:rotation} first:
\begin{equation}
  W_{-U\vb{G},\,-U\vb{G}'}(-U\vb{q}_\text{IBZ})
  = W_{-\vb{G},\,-\vb{G}'}(-\vb{q}_\text{IBZ}),
\end{equation}
and then the reality condition Eq.~\eqref{eq:reality}:
\begin{equation}\label{eq:conjugation}
  W_{-\vb{G},\,-\vb{G}'}(-\vb{q}_\text{IBZ})
  = W_{\vb{GG}'}(\vb{q}_\text{IBZ})^*.
\end{equation}
Hence, when $s = -1$, the stored $W_{\vb{GG}'}(\vb{q}_\text{IBZ})$
must be \textbf{complex-conjugated} before use.

%----------------------------------------------------------------------
\section{Why $W_{\vb{GG}'}(\vb{q})$ is complex for non-centrosymmetric
         crystals}
\label{sec:imW}
%----------------------------------------------------------------------

Since the conjugation has no effect on a real matrix, the bug is only
relevant when $W_{\vb{GG}'}(\vb{q})$ has a nonzero imaginary part.
In this section we show that this imaginary part is a genuine physical
property of non-centrosymmetric crystals, not a numerical artifact.

\subsection{Reality in real space does not imply reality in reciprocal
            space}

The static screened interaction $W(\vb{r}, \vb{r}', \omega = 0)$ is
real for all $\vb{r}$, $\vb{r}'$.  However, reality in real space and
reality in reciprocal space are not simultaneously achievable unless the
crystal has inversion symmetry.  This is a standard property of Fourier
transforms, analogous to the fact that the structure factor $S(\vb{G})$
of a non-centrosymmetric crystal is complex even though the electron
density $n(\vb{r})$ is real.

\subsection{Centrosymmetric case: $W_{\vb{GG}'}$ is real}

For a crystal with spatial inversion, $W(\vb{r},\vb{r}') =
W(-\vb{r},-\vb{r}')$.  Taking the complex conjugate of the Fourier
transform and substituting $\vb{r} \to -\vb{r}$,
$\vb{r}' \to -\vb{r}'$:
\begin{align}
  W_{\vb{GG}'}(\vb{q})^*
  &= \frac{1}{\Omega}\int\!\!\int
     e^{i(\vb{q}+\vb{G})\cdot\vb{r}}\,
     W(\vb{r},\vb{r}')\,
     e^{-i(\vb{q}+\vb{G}')\cdot\vb{r}'}\,
     \dd{\vb{r}}\,\dd{\vb{r}'}
  \notag\\
  &= \frac{1}{\Omega}\int\!\!\int
     e^{-i(\vb{q}+\vb{G})\cdot\vb{r}}\,
     W(-\vb{r},-\vb{r}')\,
     e^{i(\vb{q}+\vb{G}')\cdot\vb{r}'}\,
     \dd{\vb{r}}\,\dd{\vb{r}'}
  \notag\\
  &= \frac{1}{\Omega}\int\!\!\int
     e^{-i(\vb{q}+\vb{G})\cdot\vb{r}}\,
     W(\vb{r},\vb{r}')\,
     e^{i(\vb{q}+\vb{G}')\cdot\vb{r}'}\,
     \dd{\vb{r}}\,\dd{\vb{r}'}
  \notag\\
  &= W_{\vb{GG}'}(\vb{q}).
\end{align}
Hence $W_{\vb{GG}'}(\vb{q})$ is real for \emph{all} $\vb{G}$,
$\vb{G}'$, and the bug has no effect.

\subsection{Non-centrosymmetric case: $W_{\vb{GG}'}$ is genuinely
            complex}

Without inversion symmetry, the substitution
$W(-\vb{r},-\vb{r}') = W(\vb{r},\vb{r}')$ is not valid, so
$W_{\vb{GG}'}(\vb{q})^* \neq W_{\vb{GG}'}(\vb{q})$ in general.
The imaginary part is a real physical property of the crystal,
entirely independent of any numerical broadening parameter~$\eta$.

We can verify this at the level of the irreducible polarizability.  In
GPAW, the static $\chi^0$ is computed as
\begin{equation}\label{eq:chi0}
  \chi^0_{\vb{GG}'}(\vb{q}, \omega = 0) = \sum_{nn'\vb{k}}
  \frac{f_{n\vb{k}} - f_{n'\vb{k}+\vb{q}}}
       {\varepsilon_{n\vb{k}} - \varepsilon_{n'\vb{k}+\vb{q}} + i\eta}
  \;\rho^*_{nn'\vb{G}}(\vb{k},\vb{q})\;
  \rho_{nn'\vb{G}'}(\vb{k},\vb{q}),
\end{equation}
where
$\rho_{nn'\vb{G}}(\vb{k},\vb{q})
= \langle \psi_{n\vb{k}} | e^{i(\vb{q}+\vb{G})\cdot\vb{r}} |
  \psi_{n'\vb{k}+\vb{q}} \rangle$
are pair densities.  In the code (class \texttt{GenericUpdate} in
\texttt{gpaw/response/integrators.py}), the actual formula evaluated at
$\omega = 0$ is
\begin{equation}\label{eq:xm}
  x_m = \frac{1}{0 + \Delta\varepsilon_m + i\eta}
       - \frac{1}{0 - \Delta\varepsilon_m + i\eta}
       = \frac{2\Delta\varepsilon_m}{\Delta\varepsilon_m^2 + \eta^2},
\end{equation}
which is \emph{real}.  The imaginary part of $\chi^0$ therefore does
\emph{not} come from the energy denominator (which cancels at $\omega =
0$ regardless of $\eta$).  It comes from the pair-density outer product
$\rho^*_{nn'\vb{G}}\,\rho_{nn'\vb{G}'}$, which is a complex matrix
whenever $\vb{G} \neq \vb{G}'$ (it equals $|\rho|^2 \in \mathbb{R}$
only on the diagonal).

With inversion symmetry, Bloch states can be chosen such that the pair
densities satisfy
$\rho_{nn'\vb{G}}(\vb{k},\vb{q})
= \rho^*_{nn',-\vb{G}}(\vb{k},\vb{q})$,
which makes the full $\chi^0_{\vb{GG}'}$ real.  Without inversion
symmetry, no such gauge choice is possible, and
$\operatorname{Im}\chi^0_{\vb{GG}'} \neq 0$, propagating through
$\varepsilon^{-1}$ to give a genuinely complex $W_{\vb{GG}'}$.

%----------------------------------------------------------------------
\section{The error introduced by the missing conjugation}
%----------------------------------------------------------------------

The error per matrix element is
\begin{equation}
  W_{\vb{GG}'}(\vb{q}) - W_{\vb{GG}'}(\vb{q})^*
  = 2i\,\operatorname{Im} W_{\vb{GG}'}(\vb{q}).
\end{equation}
While the imaginary parts are nonzero, they are typically much smaller
in magnitude than the real parts (recall that $W$ is dominated by the
real bare Coulomb interaction).  This explains why the error is modest
($\sim 10^{-5}$\,Ha per BSE matrix element), while still being large
enough to break the Hermiticity of the BSE Hamiltonian.

%----------------------------------------------------------------------
\section{The bug: missing conjugation in the BSE kernel}
%----------------------------------------------------------------------

The direct kernel is assembled in \texttt{add\_direct\_kernel}.  After
computing the pair densities $\rho^{(3)}_{nn'\vb{G}}$ (valence) and
$\rho^{(4)}_{nn'\vb{G}}$ (conduction), the kernel matrix element is
\begin{equation}
  W_{mm'MM'} = \sum_{\vb{G},\vb{G}'}
    \rho^{(3)*}_{mm'\vb{G}}\;
    W_{\vb{GG}'}\;
    \rho^{(4)}_{MM'\vb{G}'}.
\end{equation}
In the code:

\begin{lstlisting}[style=python,firstnumber=702,
  caption={\texttt{gpaw/response/bse.py}: Contraction of pair densities
           with $W$ in \texttt{add\_direct\_kernel}.}]
W_mmmm = np.einsum(
    'ijk,km,pqm->ipjq',
    rho3_nnG.conj(),
    W_GG,
    rho4_nnG,
    optimize='optimal')
\end{lstlisting}

The matrix \verb|W_GG| was taken directly from the IBZ storage with no
conjugation applied when $s = -1$.  The fix adds a single conjugation:

\begin{lstlisting}[style=python,firstnumber=698,
  caption={\texttt{gpaw/response/bse.py}: The fix --- conjugate $W$
           when the symmetry operation includes time reversal.}]
if symop.sign == -1:
    W_GG = W_GG.conj()
\end{lstlisting}

%----------------------------------------------------------------------
\section{Why GW is unaffected}
%----------------------------------------------------------------------

The GW self-energy involves a \emph{quadratic form}:
\begin{equation}
  \Sigma \;\sim\; \sum_{\vb{G},\vb{G}'}
    \rho^*_{\vb{G}}\, W_{\vb{GG}'}\, \rho_{\vb{G}'},
\end{equation}
with the \emph{same} pair density $\rho$ on both sides.  For Hermitian
$W$:
\begin{equation}
  \qty(\rho^\dagger W \rho)^*
  = \sum_{\vb{G},\vb{G}'}
    \rho_{\vb{G}}\, W^*_{\vb{GG}'}\, \rho^*_{\vb{G}'}
  = \sum_{\vb{G},\vb{G}'}
    \rho^*_{\vb{G}'}\, W_{\vb{G}'\vb{G}}\, \rho_{\vb{G}}
  = \rho^\dagger W \rho.
\end{equation}
The result is always real, so replacing $W$ by $W^*$ yields the same
value.  The BSE kernel, by contrast, uses a \emph{bilinear form}
$\rho^{(3)\dagger} W\, \rho^{(4)}$ with two different pair densities, so
the conjugation of $W$ produces a genuinely different result.

\end{document}
